Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-11-07 _tex/index.tex
Built site for gh-pages

In \emph{comparative judgment} (CJ) studies, judges assess the presence
of a trait or competence by conducting pairwise comparisons of stimuli
\citep{Thurstone_1927, Pollitt_2004, Pollitt_2012a}. The comparison
produces a dichotomous outcome, indicating which stimulus is perceived
to possess a higher trait level. After conducting multiple rounds of
pairwise comparisons, researchers use the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959} to process the outcomes and
estimate scores that reflect the underlying trait of interest. This
method has been successfully employed in assessing the quality of
written texts, where quality describes the underlying trait of interest
and the texts serve as the stimuli
\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.

@@

Researchers then typically use the estimated BTL scores, or their
transformations, to conduct additional analyses or hypothesis testing.
The scores have been used to identify `misfit' judges and stimuli
\citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018}, detect
biases in judges' ratings \citep{Pollitt_et_al_2003, Pollitt_2012b},
calculate correlations with other assessment methods
\citep{Goossens_et_al_2018, Bouwer_et_al_2023}, or test hypotheses
related to the underlying trait of interest
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.

However, the statistical literature cautions against using estimated
scores to conduct additional analyses or tests. A key consideration is
that BTL scores are parameter estimates that inherently carry
uncertainty. Ignoring this uncertainty when conducting separate analyses
and tests can inflate their precision and statistical power, increasing
the risk of committing a type I error \citep{McElreath_2020}. A type I
error results when a null hypothesis is incorrectly rejected
\citep{Everitt_et_al_2010}.

To mitigate these risks, principles from Structural Equation Modeling
(SEM) \citep{Hoyle_et_al_2023, Kline_et_al_2023} and Item Response
Theory (IRT) \citep{deAyala_2009, Fox_2010, vanderLinden_et_al_2017}
recommend conducting these analyses and tests within a structural model.
A structural model specify how different manifest or latent variables
influence the latent variable of interest \citep{Everitt_et_al_2010}.
This approach allows analyses that can account for both the scores and
their uncertainties simultaneously, rather than treating them as
separate elements. Therefore, an integrated approach that combines CJ's
structural and measurement models can offer significant advantages.

@@

From early on in the literature, it has been clear that the BTL model
represents a statistical articulation of Thurstone's Case 5
\citeyearpar{Thurstone_1927}. Talk about \citet{Pollitt_et_al_2003} and
\citet{Bramley_2008}.

What case 5 implies, Assumptions. Not a normal distribution but a
logistic distribution

Table with assumptions

@@

\subsection{Appendix A: Ignoring uncertainty}\label{sec-appendix-A}

\subsection{Appendix B: The five cases of
Thurstone}\label{sec-appendix-B}

>>>>

In \emph{comparative judgment} (CJ) studies, judges assess the presence
of a trait or competence in different stimuli by performing pairwise
comparisons \citep{Thurstone_1927, Pollitt_2004, Pollitt_2012a}. The
comparison produces a dichotomous outcome, indicating which stimulus is
perceived to possess a higher trait level. After multiple rounds of
pairwise comparisons, researchers use the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959} to process the resulting outcomes
and estimate scores that reflect the underlying trait of interest. This
method has been successfully employed in assessing the quality of
written texts, where quality describes the underlying trait of interest
and the texts serve as the stimuli
\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.

@@

Researchers then typically use the estimated BTL scores, or their
transformations, to conduct additional analyses or hypothesis tests. For
example, these scores have been used to identify `misfit' judges and
stimuli \citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018},
detect biases in judges' ratings
\citep{Pollitt_et_al_2003, Pollitt_2012b}, calculate correlations with
other assessment methods \citep{Goossens_et_al_2018, Bouwer_et_al_2023},
or test hypotheses related to the underlying trait of interest
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.

However, the statistical literature advises caution when using estimated
scores to conduct additional analyses or hypotheses tests. A key
consideration is that BTL scores are parameter estimates that inherently
carry uncertainty. Ignoring this uncertainty can introduce bias into the
analysis and reduce the precision of hypothesis tests. Notably, the
direction and magnitude of this bias are often unpredictable; results
may be attenuated, exaggerated, or remain unaffected, depending on the
uncertainty in the scores and the actual effects being tested
\citetext{\citealp[p.~25]{Kline_et_al_2023}; \citealp[p.~137]{Hoyle_et_al_2023}}.
Furthermore, reduced precision in hypothesis tests weakens their
statistical power, ultimately increasing the likelihood of committing
type-I or type-II errors \citep{McElreath_2020}.

To mitigate these risks, principles from Structural Equation Modeling
(SEM) \citep[p.~138]{Kline_et_al_2023, Hoyle_et_al_2023} and Item
Response Theory (IRT)
\citep{deAyala_2009, Fox_2010, vanderLinden_et_al_2017} recommend
conducting these analyses and tests within a structural model. A
structural model specifies how different manifest or latent variables
influence the latent variable of interest \citep{Everitt_et_al_2010}.
This approach allows analyses that can account for both the scores and
their uncertainties simultaneously, rather than treating them as
separate elements. Therefore, an integrated approach that combines CJ's
measurement and structural models can offer significant advantages.

@@

@@
